rawPaper

### 2020-10-30 analysis/paper/rawPaper.Rmd
modelTestig here we go

                                                      # INSTALLING PACKAGES AND LIBRARIES - if needed
packages = c("rentrez", "ape", "seqinr", "BiocManager", "msa")
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)

#if (!requireNamespace("BiocManager", quietly=TRUE))
#install.packages("BiocManager")
#BiocManager::install("msa")

                                                       # DOWNLOADING SEQS FROM GENBANK

@@

library(ape)   # delete this line after finishing

@@

                                     # ALIGNIG/EDITTING SEQS AND GETTING THEM ALL READY FOR THE PHYLOGENETIC ANALYSES

### Removing sequences of incomplete/unspecific NAMES; that is, those with 'sp.'
nameFasta <- read.fasta(file = "../data/raw_data/Sicydium_COI_seqinrFormat", seqtype = "DNA", as.string = TRUE, forceDNAtolower = FALSE)

fastaNames <- strsplit(names(nameFasta), '_')                  # creating a vector splitting species names
completeSeqs <- which(! grepl("sp.", fastaNames))              # selecting only those with specific name in them       # 562

@@

### Removing sequences of LENGTH shorter than 600 pb, given that the gene COI has between 627 and 658 bp.
lengthFasta <- read.fasta(file = "../data/raw_data/Sicydium_COI_names.fasta", seqtype = "DNA", as.string = TRUE, forceDNAtolower = FALSE)
fastaLength <- lengthFasta[which(getLength(lengthFasta) >= 600)]       # 399
write.fasta(sequences = fastaLength, names = names(fastaLength), file.out = "../data/raw_data/Sicydium_COI_lengths.fasta")

### Aligning the sequences
# using libraries "msa" and "Biostrings"

library(msa)           # delete this line after finishing
library(Biostrings)    # delete this line after finishing

mySequences <- readDNAStringSet("../data/raw_data/Sicydium_COI_lengths.fasta")

myClustalOmegaAlignment <- msa(mySequences, "ClustalOmega")   # using algorithm Clustal Omega (Sievers et al. 2011) for aligning
#print(myClustalOmegaAlignment, showConsensus=FALSE, showNames=TRUE, show="complete")

# removing samples with to many gaps at the beginning and the end of the sequences 
colmask(myClustalOmegaAlignment) <- IRanges(start=c(1,666),end=c(13,679))                
myClustalOmegaAlignment2 <- as(myClustalOmegaAlignment, "DNAStringSet")
writeXStringSet(myClustalOmegaAlignment2, file="../data/raw_data/Sicydium_COI_alignment.fasta")         # fasta file of 652 bp

### Removing most duplicate sequences

### Removing the segment before the start codon TAC

@@

# the alternatie is var1 = read.FASTA = (sequences, type = 'DNA')

>>>>

                                                      # INSTALLING PACKAGES AND LIBRARIES - if needed
packages = c("rentrez", "ape", "seqinr", "BiocManager", "modelTest")
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("msa")
BiocManager::install("systemPipeR")
library(stringr)
library(systemPipeR)
                                                       # DOWNLOADING SEQS FROM GENBANK

@@

# using library "ape"
library(ape)   # delete this line after finishing

@@

                                     # ALIGNIG/EDITTING SEQS AND GETTING THEM ALL READY FOR THE PHYLOGENETIC ANALYSIS

### REMOVING sequences of incomplete/unspecific NAMES; that is, those with 'sp.'
nameFasta <- read.fasta(file = "../data/raw_data/Sicydium_COI_seqinrFormat", seqtype = "DNA", as.string = TRUE, forceDNAtolower = FALSE)

fastaNames <- strsplit(names(nameFasta), '_')                  # creating a vector splitting species names
completeSeqs <- which(! grepl("sp.", fastaNames))              # selecting only those with specific name in them       # 562 seqs

@@

### REMOVING sequences of LENGTH shorter than 600 pb, given that the gene COI has between 627 and 658 bp.
lengthFasta <- read.fasta(file = "../data/raw_data/Sicydium_COI_names.fasta", seqtype = "DNA", as.string = TRUE, forceDNAtolower = FALSE)
fastaLength <- lengthFasta[which(getLength(lengthFasta) >= 600)]       
write.fasta(sequences = fastaLength, names = names(fastaLength), file.out = "../data/raw_data/Sicydium_COI_lengths.fasta")      # 399 seqs

### Aligning the sequences    # using libraries "msa" and "Biostrings"
library(msa)           # delete this line after finishing
library(Biostrings)    # delete this line after finishing

alignmentFasta <- readDNAStringSet("../data/raw_data/Sicydium_COI_lengths.fasta")
ClustalOAlignment <- msa(alignmentFasta, "ClustalOmega")   # algorithm Clustal Omega (Sievers et al. 2011) used for aligning
#print(ClustalOAlignment, showConsensus=FALSE, showNames=TRUE, show="complete")

# REMOVING samples with to many GAPS at the beginning and the end of the sequences 
colmask(ClustalOAlignment) <- IRanges(start=c(1,666),end=c(13,679))                
ClustalOAlignment2 <- as(ClustalOAlignment, "DNAStringSet")
writeXStringSet(ClustalOAlignment2, file="../data/raw_data/Sicydium_COI_alignment.fasta")         # 399 seqs of 652 bp.

### Removing DUPLICATE sequences    # using "seqinr" to do it

# for not using attributes to name sequences
duplicatesFasta <- read.fasta("../data/raw_data/Sicydium_COI_alignment.fasta", set.attributes = FALSE) 
seqnames <- names(duplicatesFasta)
dups <- grep(TRUE,duplicated(duplicatesFasta))
uniqueNames <- seqnames[-dups]
uniqueSeqs <- duplicatesFasta[-dups] 
write.fasta(sequences = uniqueSeqs, names = uniqueNames, file.out = "../data/raw_data/Sicydium_COI_dupli.fasta")      # 197 seqs of 652 bp.

### Replacing gaps with Ns
library(stringr)

gapsFasta <- read.fasta(file = "../data/raw_data/Sicydium_COI_alignment.fasta", seqtype = "DNA", as.string = TRUE, forceDNAtolower = FALSE)
replacingGaps <- lapply(gapsFasta, function(x) str_replace_all(x, "-", "N"))
write.fasta(replacingGaps,names(replacingGaps),"../data/raw_data/Sicydium_COI_gaps.fasta")

# GETTING the READING FRAME
BiocManager::install("systemPipeR")    # delete this line after finishing
library(systemPipeR)

gettingRF <- function(x){
  fasta <- readDNAStringSet("../data/raw_data/Sicydium_COI_gaps.fasta")
  orf1 <- predORF(fasta[1:10], n=1, type="df", mode="orf", strand="sense")   # predORF() do not read seqs with gaps
  # predicts longest ORF for sense strand in each query sequence                                                                          
  longestSeq1 <- orf$start[orf$width == max(orf$width)]
  codonPosition1 <- longestSeq[1]/3
  codonConfirmation1 <- round(codonPosition1)
  if (codonPosition1==codonConfirmation1){
    print("The Reading Frame of the seqs is +1, no further action is required, go to modelTesting")
    } else {
    if (codonPosition1!=codonConfirmation1){
      #remove first nucleotide position
      newRF2 <- subseq(fasta, start = 2, end = tempLength)      # 651 bp

      # getting new RF from the new alignment with -1 bp.
      orf2 <- predORF(newRF2[1:10], n=1, type="df", mode="orf", strand="sense")

      # checking size gain
      longestSeq2 <- orf2$start[orf2$width == max(orf2$width)]
      codonPosition2 <- (longestSeq2[1]+2)/3
      codonConfirmation2 <- round(codonPosition2)
      if (codonPosition2==codonConfirmation2){
        writeXStringSet(newRF2, file="../data/raw_data/Sicydium_COI_newRF2.fasta")
      print("The Reading Frame of the seqs is +2, new fasta file created, go to modelTesting")
      } else {
      if (codonPosition2!=codonConfirmation2){
        #remove first nucleotide position
        tempLength <- length(fasta[[1]])
        newRF3 <- subseq(fasta, start = 3, end = tempLength)
        orf3 <- predORF(newRF3[1:10], n=1, type="df", mode="orf", strand="sense") 
        # checking size gain
        longestSeq3 <- orf3$start[orf3$width == max(orf3$width)]
        codonPosition3 <- (longestSeq3[1]+2)/3
        codonConfirmation3 <- round(codonPosition3)
        if (codonPosition3==codonConfirmation3){
          writeXStringSet(newRF3, file="../data/raw_data/Sicydium_COI_newRF3.fasta")
        print("The Reading Frame of the seqs is +3, new fasta file created, go to modelTesting")
       }
     }
   }  
    }
  }
}

gettingRF(x)

@@

# the alternative is var1 = read.FASTA = (sequences, type = 'DNA')
